High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2021-04-06 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

The ring width can be measured in polar coordinates by fitting 1-D Gaussians, thus obtaining a width and centroid at each azimuth (see Fig.~\ref{fig:polarring}). On average, we obtained a FWHM of 2.83$\pm$0.50\,au, and a stellocentric radius of 13.15$\pm$0.26\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim0\farcs021 \,\times \,\sim0\farcs018$, we see that Ring13 is marginally resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.56\,au.

In a second optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,$\degr$, with an inclination of 33.92$\pm$0.01\,$\degr$ and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 share a very similar orientation and centre, given the errors. However, perhaps due to the joint effect of all these small differences, we can see that there are some hints for a somewhat different orientation, as summarised in Fig.~\ref{fig:polarring}.

Interestingly, the ALMA image also detects faint 1.3\,mm continuum emission near the stellar positions (see the inset in Fig.~\ref{fig:images_vs_simulated}). Since this faint central emission is larger than the angular resolution, it is probably stemming from thermal emission from  large dust grains rather than directly from the stars. The main blob of this dust structure is at estimated distance of only $0\farcs012\pm0\farcs002$, or $\sim$0.85$\pm$0.14\,au from the binary system, and, estimating its form as a Gaussian ellipse, it has a mean FWHM of $\sim$1.43\,au. If this feature is caused by the presence of millimetre-sized dust, assuming optically thin emission, its estimated mass would be around 0.12\,M$_{\earth}$. %and may be part of an inner disc introduced in the RT model to match the SED (see Sec.~\ref{sec:model} below).

@@

\subsection{Spectral energy distribution}

The observed SED was collected from data in the literature \citep{1988iras....7.....H, 1990A&A...234..230H, Jensen_97, 2000A&A...355L..27H, 2001KFNT...17..409K, 2003yCat.2246....0C, 2007PASJ...59S.369M, 2008PASP..120.1128O, 2010A&A...514A...1I, 2012yCat.2311....0C}, available online in \textsc{vizier}. We also  used  archival \textit{Spitzer} IRS spectroscopic data available in the CASSIS database \citep{Lebouteiller_2015} (see Fig.~\ref{fig:SED}).

The SED features the standard signature of a transition disk, with a distinctive “dip” in the continuum near 10 $\micron$m suggesting that small, warm dust is preferentially depleted (although not absent; see Section 4.2.1) near the central binary. although the very weak excess
in the near-infrared makes it clear that the dust optical depths
must be significantly diminished within at least a few AU of the
central stars*** \citep{Rosenfeld_2013}.

@@

The parametric scale height profiles for the gas and for each dust population are 
\begin{equation}
    \label{scale}
  H(r)=\chi \, H_{\circ} \,\left( \frac{r}{r_{\circ}}\right)^{1+\psi(r)},
\end{equation}
where $H_\circ$ is the scale height at $r$ = $r_\circ$, $\psi$ is the flaring index and $\chi$ is a scaling factor (in the range $0-1$) that mimics dust settling. The gas and the small dust (sd) grains have no settling, so $\chi_{\mathrm{gas}}$ = $\chi_{\mathrm{sd}}$ = 1. The radial spread of Ring13 appears two to three times wider in the gas-tracing NIR observations than in the fluxes received from larger grains with ALMA. We assume that the same ratio holds in the vertical direction (due to the settling of larger grains towards the midplane of the disc), leading to $\chi_{\mathrm{ld}}=0.4$. This is analogous to assuming equal radial and vertical turbulent diffusions.

@@

The inner disc follows a power-law function defined by
\begin{equation}
  \Sigma_{\mathrm{inner\,disc}}(r) =\Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \mathrm{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density power-law index. We used $R_c$ = 16\,au, $\Sigma_\mathrm{c} =3.3\times10^{-4}$\,$\mathrm{g\,cm^{-3}}$ and a fixed $\gamma$ = 1 as it is a typical value for discs \citep{Andrews_2009,Andrews_2010}. Our model extends from 0.3\,au outwards, as the interior is expected to be cleared by dynamical effects of the central binary \citep{Art_Lu}

@@

The simulated DPI image at 1.65\,$\micron$ in Fig.~\ref{fig:images_vs_simulated} was obtained with the scattering matrix calculated by the {\tt makeopac.py} script provided in the \textsc{radmc3d} package. As a way of reproducing the \textit{H}-band image, we used a different grain size distribution, where we centred a Gaussian at 0.4\,$\micron$, smeared out by 30\% ($\sigma$ = 0.12\,$\micron$), and with distributed over 20 bins within that range. This distribution applies only to generate the NIR $Q_\phi$ image. To produce this $Q_\phi$ image we performed a linear combination of the two orthogonal linear polarizations \textit{U} and \textit{Q}, following \citet{Avenhaus_2017}, which gives a representation of an unbiased estimate of the polarized intensity image.

@@

While certainly not unique, our parametric model is fairly successful in reproducing the available data. The simulated images and the SED of the model are shown in Fig.~\ref{fig:images_vs_simulated} and Fig.~\ref{fig:SED} respectively.

@@

The effective beam size of the presented ALMA observations marginally resolves the FWHM of Ring13 to be $w_\mathrm{d} = 2.46\pm$0.56\,au. Using that the gas' Ring13-FWHM in the RT model corresponds to $w_\mathrm{g} = 5.42\mathrm{au}$, we can see that large dust grains have to be radially converged to explain the discrepancy.

Even though the radial spread of large dust grains in Ring13 appears to be quite thin, the width in comparison to the sub-lying gas profile speaks for the presence of considerable turbulent diffusion. Following a similar ring analysis as in \citet{2018ApJ...869L..46D}, we find that the ratio between the dimensionless diffusion parameter, $D$, and the dimensionless Stokes number, St, (which parameterizes the dynamical behaviour of a grain) is roughly, $D/{\rm St}\approx 0.1$. The observed signal is expected to be dominated by grains of size $a\approx 0.02\,{\rm cm}$. The RT model, together with the dust-to-gas ratio of 0.05, prescribe a gas density of $\Sigma_{\rm g} \approx 0.5\,{\rm g}{\rm cm}^{-2}$ to the location of Ring13. With these values, the relevant Stokes number is approximated to be St$\approx 0.1$. This yields an estimate for the level of diffusivity of $D\approx 0.01$. It further provides a value for the level of turbulent viscosity in Ring13, $\alpha_{\rm turb}\approx 0.01$, assuming viscosity to be equal to the level of diffusion \citep{2007Icar..192..588Y}. We note that an observation of molecular line broadening has found no evidence for turbulent contributions, suggesting $\alpha_{\rm turb}<0.01$ \citep{Flaherty_2020}. The value inferred from our model is just within this limit. It is linearly proportional to the local dust-to-gas ratio (by our definition of the gas surface density profile) a standard ISM  value of 0.01 for the dust-to-gas ratio would thus lead to an expected level of turbulent viscosity of $\alpha_{\rm turb}\approx 5\times10^{-3}$. While the exact value for $\alpha_{\rm turb}$ is not well constrained, the presence of turbulence is a robust detection, as it is required to explain the radial spread of the resolved Ring13.

>>>>

The ring width can be measured in polar coordinates by fitting 1-D Gaussians at each azimuth, thus obtaining a width and centroid. In Fig.~\ref{fig:polarring}) is shown the positions of the obtained centroids and widths over the polar decomposition of the ALMA image. On average, we obtained a FWHM of 2.83$\pm$0.50\,au, and a stellocentric radius of 13.15$\pm$0.26\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim0\farcs021 \,\times \,\sim0\farcs018$, we see that Ring13 is marginally resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.56\,au.

In a second optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,$\degr$, with an inclination of 33.92$\pm$0.01\,$\degr$ and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 share a very similar orientation and centre, given the errors. However, very likely due to the joint effect of all these small differences, we can see that there are some hints for a somewhat different orientation. This is summarised in the second panel of Fig.~\ref{fig:polarring}, where is showing the difference between the Gaussian fitting at each azimuth for both optimization. Both fitted centroids follow a curve almost entirely within the errors of the other one, leaving the window open to the discussion of if this differences might be just a consequence of the pointing error or are they a real notable feature of the disc.

Interestingly, the ALMA image also detects faint 1.3\,mm continuum emission near the stellar positions (see the inset in Fig.~\ref{fig:images_vs_simulated}). Since this faint central emission is larger than the angular resolution, it is probably stemming from thermal emission from  large dust grains rather than directly from the stars. The main blob of this dust structure is at estimated distance of only $0\farcs012\pm0\farcs002$, or $\sim$0.85$\pm$0.14\,au from the binary system, and, estimating its form as a Gaussian ellipse, it has a mean FWHM of $\sim$1.43\,au. If this feature is caused by the presence of millimetre-sized dust, assuming optically thin emission, i.e. following
\begin{multline}
    \label{optically_thin_equation}
      M_{blob} = 0.03 M_{\sun} \left(\frac{F}{1\,\mathrm{Jy}}\right) \left(\frac{d}{100\,\mathrm{pc}}\right)^2 \left(\frac{\lambda}{1.3\,\mathrm{mm}}\right)^3 
      \\
      \left(\frac{50\,\mathrm{K}}{T}\right)    \left(\frac{0.02\,\mathrm{cm^2/g}}{\kappa_{1.3\,\mathrm{mm}}}\right),
\end{multline}
where $F$ is the total flux of the blob, $d$ the distance of 71.4\,pc , $\lambda$ the wavelength, $\kappa_{1.3\,\mathrm{mm}}$ the opacity at 1.3\,mm and the temperature set at $T=500\,\mathrm{K}$. Then, its estimated mass would be around 0.12\,M$_{\earth}$.

@@

\subsection{Spectral energy distribution} \label{subsec:SED}

The observed SED was collected from data in the literature \citep{1988iras....7.....H, 1990A&A...234..230H, Jensen_97, 2000A&A...355L..27H, 2001KFNT...17..409K, 2003yCat.2246....0C, 2007PASJ...59S.369M, 2008PASP..120.1128O, 2010A&A...514A...1I, 2012yCat.2311....0C}, available online in \textsc{vizier}. We also  used  archival \textit{Spitzer} IRS spectroscopic data available in the CASSIS database \citep{Lebouteiller_2015} (see Fig.~\ref{fig:SED}). The SED describe the characteristic “dip” near 10 $\micron$, a common element of transition disk, as described by \citet{Rosenfeld_2013}. \citet{Jensen_97} concluded that these data matched that of a extended circumbinary disk truncated at $\sim$0.3\,au, as the interior would also be expected to be cleared by dynamical effects of the central binary \citep{Art_Lu}.

@@

The parametric scale height profiles for the gas and for each dust population are 
\begin{equation}
    \label{scale}
  H(r)=\chi \, H_{\circ} \,\left( \frac{r}{r_{\circ}}\right)^{1+\psi(r)},
\end{equation}
where $H_\circ$ is the scale height at $r$ = $r_\circ$, $\psi$ is the flaring index and $\chi$ is a scaling factor (in the range $0-1$) that mimics dust settling. In hydrostatical equilibrium dust diffusion and settling are expected to balance each other, leading to a settling factor of $\chi = \sqrt{D_{\rm d}/(D_{\rm d} + St)}$ \citep{Dubrulle1995}. Here, $D_{\rm d}$ is a dimensionless parameter that informs about the level of diffusivity (it is typically assumed to be similar to the level of turbulence for the particle sizes regarded here \citep{2007Icar..192..588Y}. The Stokes number,
\begin{equation}
    {\rm St} := 0.5 \,\pi a \rho_{\rm mat} \, \Sigma_{\rm gas}^{-1}, 
\end{equation}
summarises a particle's dynamical behaviour in a given environment (where $\rho_{\rm mat}$ represents a dust particle's material density).. By definition, the gas has no settling, and the small dust grain's settling is negligible. As we do not know the level of diffusivity in the disc (this will be a topic of interest in Section~\ref{sec:results}), nor the exact value of $\Sigma_{\rm gas}$, we infer the scaling factor for large dust, $\chi_{\rm ld}$, from the width of Ring13. In the radial profile this ring is observed two to three times wider in the gas-tracing NIR than in the fluxes received from larger grains by ALMA. The gas and the small dust (sd) grains have no settling, so $\chi_{\mathrm{gas}}$ = $\chi_{\mathrm{sd}}$ = 1. The radial spread of Ring13 appears two to three times wider in the gas-tracing NIR observations than in the fluxes received from larger grains with ALMA. We assume that the same ratio holds in the vertical direction (due to the settling of larger grains towards the midplane of the disc), leading to $\chi_{\mathrm{ld}}=0.4$. This is analogous to assuming equal radial and vertical turbulent diffusions.

@@

The inner disc follows a power-law function defined by
\begin{equation}
  \Sigma_{\mathrm{inner\,disc}}(r) =\Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \mathrm{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density power-law index. We used $R_c$ = 16\,au, $\Sigma_\mathrm{c} =3.3\times10^{-4}$\,$\mathrm{g\,cm^{-3}}$ and a fixed $\gamma$ = 1 as it is a typical value for discs \citep{Andrews_2009,Andrews_2010}. Our model extends from 0.3\,au outwards, consistent with the inner edge radius inferred from the SED data (See sec. \ref{subsec:SED}).

@@

The simulated DPI image at 1.65\,$\micron$ in Fig.~\ref{fig:images_vs_simulated} was obtained with the scattering matrix calculated by the {\tt makeopac.py} script provided in the \textsc{radmc3d} package. As a way of reproducing the \textit{H}-band image, we used a different grain size distribution, where we centred a Gaussian at $\mathrm{a}$ = 0.4\,$\micron$ with $\sigma_{\mathrm{a}}$ = 0.12\,$\micron$ (smeared out by 30\%), and distributed the dust over 20 bins within the range of $\pm \sigma$. This distribution applies only to generate the NIR $Q_\phi$ image and not the ALMA image or the SED. To produce this $Q_\phi$ image we performed a linear combination of the two orthogonal linear polarizations \textit{U} and \textit{Q}, following \citet{Avenhaus_2017}, which gives a representation of an unbiased estimate of the polarized intensity image.

@@

While certainly not unique, our parametric model is fairly successful in reproducing the available data. The simulated images and the SED of the model are shown in Fig.~\ref{fig:images_vs_simulated} and Fig.~\ref{fig:SED} respectively. The model presented here is a solution of a very degenerative problem, ergo, as a way of computing a measure of goodness of our model, we calculated a chi--square optimization to find the uncertainties of the some relevant parameters that will be useful for the discussion. Thus, we found that the width of Ring13 is $w_\mathrm{g} = 5.42^{+0.01}_{-1.15}$\,au, the scale height is $H_\circ = 2.2^{+0.01}_{-0.03}$\,au at $r_\circ = 18$\,au, the location of Ring6 is.

@@

The effective beam size of the presented ALMA observations marginally resolves the FWHM of Ring13 to be $w_\mathrm{d} = 2.46\pm0.56$\,au. Using that the gas' Ring13-FWHM in the RT model corresponds to $w_\mathrm{g} = 5.42^{+0.01}_{-1.15}$\,au, we can see that large dust grains have to be radially converged to explain the discrepancy.

Even though the radial spread of large dust grains in Ring13 appears to be quite thin, the width in comparison to the sub-lying gas profile speaks for the presence of considerable turbulent diffusion. Following a similar ring analysis as in \citet{2018ApJ...869L..46D}, we find that the ratio between the dimensionless diffusion parameter, $D_{\rm d}$, and the dimensionless Stokes number, St, (which parameterizes the dynamical behaviour of a grain) is roughly, $D_{\rm d}/{\rm St}\approx 0.1$. The observed signal is expected to be dominated by grains of size $a\approx 0.02\,{\rm cm}$. The RT model, together with the dust-to-gas ratio of 0.05, prescribe a gas density of $\Sigma_{\rm g} \approx 0.5\,{\rm g}{\rm cm}^{-2}$ to the location of Ring13. With these values, the relevant Stokes number is approximated to be St$\approx 0.1$. This yields an estimate for the level of diffusivity of $D_{\rm d}\approx 0.01$. It further provides a value for the level of turbulent viscosity in Ring13, $\alpha_{\rm turb}\approx 0.01$, assuming the level of turbulence to be equal to the level of diffusion \citep{2007Icar..192..588Y}. We note that an observation of molecular line broadening has found no evidence for turbulent contributions, suggesting $\alpha_{\rm turb}<0.01$ \citep{Flaherty_2020}. The value inferred from our model is just within this limit. By our definition of the gas surface density profile, the value inferred for the level of turbulence is linearly proportional to the local dust-to-gas ratio. Lower values than the chosen ratio of 0.05 would, therefore, lead to an equally lower level of turbulence in the assessment. While the exact value for $\alpha_{\rm turb}$ is not well constrained, a certain level of turbulence is required to explain the radial spread of the resolved Ring13.
